---
editor_options: 
  markdown: 
    wrap: 88
bibliography: references.bib
---

# Simple Linear Regression {#chapter-3}

## Description {#ch3-description}

An approach for predicting a dependent (response) value $Y$ based on a single
independent (predictor) value $X$, given that there is some proportional linear
relationship between $X$ and $Y$. That is, assuming that one unit of change in $X$
results in a consistent change (increase or decrease) in $Y$.

$$Y \approx \beta_0 + \beta_1X$$

In this equation, $\beta_0$ represents a constant offset (the *intercept*) and $\beta_1$
represents the amount $Y$ changes for each change in $X$ (the *slope*). In a **Simple
Linear Regression** model, the goal is to estimate $\hat\beta_0$ and $\hat\beta_1$
within the observed range of $X$ and $Y$ values such that the predicted values of $Y$
are as close to the actual $Y$ values as possible, generally determined by minimizing
the *Residual Sum of Squares*.[^1]

[^1]: The sum of the squares of all the residuals in a given range, such that
    $RSS = e_1^2 + e_2^2 + … + e_n^2$.

## How it Works {#ch3-how-it-works}

Given that the goal is to estimate the *slope* and *intercept* of the linear formula
representation and that we have a series of $X/Y$ observations, the values of $\beta_0$
and $\beta_1$ can be estimated via

$$\hat\beta_1 = \frac{\sum_{i=1}^n(x_i - \bar{x})(y_i-\bar{y})}{\sum_{i=1}^n(x_i-\bar{x})^2}$$
$$\hat\beta_0 = \bar{y} - \hat\beta_1\bar{x}$$ In code, this looks like:

```{r ch3-1}
x <- mtcars$hp  # horsepower
y <- mtcars$mpg # fuel efficiency, in miles/gallon

mean_diff <- \(x) x - mean(x)
slope     <- sum(mean_diff(x) * mean_diff(y))/sum(mean_diff(x)^2)
intercept <- mean(y) - slope * mean(x)

# Visualization. The line demonstrates the simple 
# linear regression fit as calculated above.
ggplot(mtcars, aes(hp, mpg)) +
  geom_point(color = "blue") +
  geom_abline(slope = slope, intercept = intercept, color = "orange") +
  annotate(
    "text", x = 300, y = 12.5, 
    label = glue("Slope: {round(slope, 2)}\nIntercept: {round(intercept, 2)}")
  ) +
  labs(
    title = "Simple Linear Regression",
    x = "Horsepower",
    y = "Miles/Gallon"
  ) +
  theme_minimal()
```

So, for a single predictor variable, *slope* and *intercept* can be estimated by a
*relatively* straightforward algorithm.

## Evaluating Validity {#ch3-evaluating-validity}

It is possible to estimate the *Standard Error*[^2]
([[Definitions#\^72c250\|definition]]) of a model using the formulae below:

[^2]: The standard deviation of its sampling distribution or an estimate of that
    standard deviation. Accounts for one standard deviation to either side of the mean
    value.

$$e_1 = y_1 - \hat\beta_0 - \hat\beta_1x_1$$ $$RSS = e_1^2 + e_2^2 + ... + e_n^2$$
$$\sigma \approx RSE = \sqrt{RSS/(n-2)}$$
$$\widehat{SE}(\hat\beta_0)^2 = \sigma^2\left[\frac{1}{n}+\frac{\bar{x}^2}{\sum_{i=1}^n{(x_i - \bar{x})^2}}\right]$$
$$\widehat{SE}(\hat\beta_1)^2 = \frac{\sigma^2}{\sum_{i=1}^n{(x_i - \bar{x})^2}}$$

In code, that looks like:

```{r ch3-2}
predicted  <- (slope * x) + intercept
residuals  <- predicted - y
rss        <- sum(residuals^2)               # residual sum of squares
n          <- length(x)                      # number of observations
rse        <- sqrt(rss / (n_observed - 2))   # residual standard error
se_intcpt  <- sqrt(rse^2 * ((1/n) + (mean(x)^2 / sum((x - mean(x))^2))))
se_slope   <- sqrt(rse^2 / sum((x - mean(x))^2))
```

The 95% confidence interval (for the slope, in this example) can be given by:
$$\hat\beta_1 \pm 2 \times SE(\hat\beta_1)$$

From the above example, it looks like this in code:

```{r ch3-3}
min_interval  <- slope - (2 * se_slope)
max_interval  <- slope + (2 * se_slope)
(conf_interval <- c(min_interval, max_interval))
```

We find that we can say with 95% confidence that the slope lies somewhere between
`r min_interval` and `r max_interval`, which is reasonably close to our input value. Is
this enough to determine whether our $X$ and $Y$ are related? For that, we can calculate
the *t-statistic*[^3] by:

[^3]: In statistics, the t-statistic is the ratio of the departure of the estimated
    value of a parameter from its hypothesized value to its standard error. It is used
    in hypothesis testing via Student's t-test. The t-statistic is used in a t-test to
    determine whether to support or reject the null hypothesis. It is very similar to
    the z-score but with the difference that t-statistic is used when the sample size is
    small or the population standard deviation is unknown. For example, the t-statistic
    is used in estimating the population mean from a sampling distribution of sample
    means if the population standard deviation is unknown. It is also used along with
    p-value when running hypothesis tests where the p-value tells us what the odds are
    of the results to have happened.

$$t = \frac{\hat\beta_1 - 0}{SE(\hat\beta_1)}$$

to measure the number of standard deviations that $\hat\beta_1$ is away from `0`. This
*t-statistic* is further used to compute the *p-value*[^4]. In general, small *p-values*
indicate that it is unlikely to observe a substantial association between the predictor
and response due to random chance, as opposed to a real association. While it is
possible to easily calculate the *t-statistic*, finding the *p-value* generally requires
a lookup in a *t-distribution table*. Thankfully, R can handle that bit for you, as
we'll see in the `tidymodels` example for this modeling approach.

[^4]: The probability of observing a result at least as extreme as the observed result
    assuming the null hypothesis is correct. Put another way, it is the probability that
    the observed relationship between predictor(s) and response(s) could be observed as
    the result of random chance.

```{r ch3-4}
(t_statistic <- (slope - 0)/se_slope)
```

The important part here is that a low *p-value* indicates that your model fit is,
actually, applicable for the data you have applied the model to, which is ultimately the
part we are interested in.

## Evaluating Fit {#ch3-evaluating-fit}

In the previous section, we saw reference to $e$ as the difference between the observed
and predicted response value for a given predictor value. For real data that can be fit
by a linear regression, it's better to assume that the actual formula is more like
$$ Y \approx \beta_0 + \beta_1X + \epsilon$$ where $\epsilon$ represents an "error
term", that is, the accumulation of **all** the unknown variables and errors that are
affecting the relationship between $X$ and $Y$. This accounts, in a sort of
*hand-waving* way, for differences between the predicted linear relationship and
observed results. The quality of a linear regression fit, which is how well the slope
and intercept have been estimated, can be determined from the *residual standard error*
(which we calculated previously) and the $R^2$ statistic. Of the two, the $R^2$
statistic is more useful and easily understood.

### Residual Standard Error {#ch3-residual-standard-error}

The *residual standard error* (RSE), that is, the standard error for the observed
residuals from the linear regression fit, is an estimate of the standard deviation of
the error term above. It can be calculated via:

$$e_1 = y_1 - \hat\beta_0 - \hat\beta_1x_1$$ $$RSS = e_1^2 + e_2^2 + ... + e_n^2$$
$$\sigma \approx RSE = \sqrt{RSS/(n-2)}$$

In general, a smaller RSE indicates a better fit of the data. RSE, however, is an
absolute measure of the lack of fit measured in the same units as $Y$, indicating how
far off any given predicted $Y$ value may be.

### $R^2$ Statistic {#ch3-rsq-statistic}

Contrasted to RSE, $R^2$ represents the *proportion* of variance explained by the error
term and is always a value between 0 and 1. It can be calculated by:

$$TSS = \sum{(y_i - \bar{y})^2}$$ $$R^2 = \frac{TSS - RSS}{TSS} = 1 - \frac{RSS}{TSS}$$

where TSS is the *total sum of squares* of the response values and RSS is the *residual
sum of squares* (discussed above). TSS is a measure of the variability in $Y$, whereas
RSS is a measure of the variability that is left unexplained after performing the
regression. In summary, this means that $R^2$ measures the proportion of variability in
$Y$ that can be explained using $X$. Thus, an $R^2$ close to 1 indicates a linear
regression fit where most of the variability in the response is explained by the
regression.

## Example {#ch3-example}

Now, with an understanding of the underlying math and relevant statistics, we can fit a
simple linear model to the example dataset we've used thus far, that is: estimating the
fuel efficiency of a vehicle from its horsepower rating.

```{r ch3-5}
# In this example, we'll forgo the training/testing split to
# be consistent with our calculated model above.

# Pre-process the data using a simple, no-frills recipe
(simple_linear_recipe
  <- recipe(mpg ~ hp, data = mtcars))

# Setup up the model
(simple_linear_model
  <- linear_reg()
  |> set_engine("lm")
  |> set_mode("regression"))

# Bundle the recipe and model into a workflow, fit the model
(simple_linear_workflow
  <- workflow()
  |> add_recipe(simple_linear_recipe)
  |> add_model(simple_linear_model)
  |> fit(data = mtcars))

# Add predictions to training data
simple_linear_predictions <- augment(simple_linear_workflow, mtcars)

# Evaluate validity
tidy(simple_linear_workflow)
```

The `tidy` function (from the `broom` [@broom] package, also included in `tidymodels`)
provides a variety of coefficients related to the model, including the relevant terms,
their estimates, and the *p-value*. In this case, the *p-value*s for both the slope (of
the `hp` term, which is the only predictor variable) and the intercept are both very
small, indicating that a simple linear model is valid for making predictions on this
data set. But, how good is the fit?

```{r ch3-6}
# Evaluate fit
metrics(simple_linear_predictions, mpg, .pred)
```

By default, `yardstick` provides three metrics for a simple linear regression model.
`rmse` is [Root-Mean-Square
Error](https://en.wikipedia.org/wiki/Root-mean-square_deviation), `mae` is [Mean
Absolute Error](https://en.wikipedia.org/wiki/Mean_absolute_error), and `rsq` is $R^2$,
our preferred metric for determining model fit.

## Considerations {#ch3-considerations}

### Limits of Applicability {#ch3-limits-of-applicability}

Predictions made using a linear regression model are *most* accurate when made within
the ranges of the observed data used to build the model. This presents a difficulty
when, for example, attempting to predict future results. Say, you want to understand
what may happen to program participation (`Y`) if the budget for your program (`X`) is
increased beyond previous levels. The further beyond the previous maximum you increase
`X`, the less likely this model is to be accurate.

### Outliers {#ch3-outliers}

The presence of outliers, or observations with an unusual response value, in the data
can have an outsized impact on quality measures like RSE and $R^2$, while not
necessarily having an equivalent impact on the fitted model parameters. Plotting
*studentized residuals* can help reveal outliers. Typically, observations whose
studentized residuals are greater than 3 should be examined and considered for
exclusion.

### High-leverage points {#ch3-high-leverage-points}

Contrasted to outliers, high-leverage points have unusual *predictor* values. These
values can have an outsized impact on the fit of the model. They can be detected by
calculating the *leverage statistic*. If an observation has a leverage statistic that
greatly exceeds $(p+1)/n$, then it may be a high-leverage point. Plotting the
studentized residual vs the leverage statistic for each observation can be used to spot
outliers and high leverage points simultaneously.
